Measuring the double-drift illusion with hand tracking

### 2021-03-11 Manuscript.Rmd
small corrections

If $K$ is 0, then there is no illusion and the perceived direction matches the external motion direction. However, if $K$ is 1, then the internal motion has equal contribution with the external motion in determining the perceived direction. For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in each of our two experiments.

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **Left:** Individual trajectories. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity. **Top right:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading. **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path over unity slope (internal / external velocity). In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Gray dashed line: rediction of vector combination model of Equation 1 with $K=1$. Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$. Purple: best fit to our data, $K=0.72$."}
plotBoundedTracking()
```

@@

The illusion does drift continuously away from the physical path forever. It can be disrupted by a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020]. These disruptions may cause the perceived path to stay at a fixed offset, travelling parallel to the physical path, or it may return toward the veridical position before resuming the illusory direction (Fig. 3). These resets may also occur spontaneously once the accumulation has gone on too long or too far. The purpose of this experiment is to capture this spontaneous resetting and determine its source.

@@

```{r fig3, echo=F, fig.height=4.55, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets. **Left:** At some point, the perceived position may stop moving further away from its actual path. For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two. **Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate. **Bottom right:** if they occur instead after some period of time, resets should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

@@

In both kinds of trials, participants first had to move the stylus to the start position of the gabor, and then fixate a point to the left or right of the centre of the workspace. The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass. After the gabor had disappeared, participants could respond in one of two ways. When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line. Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE. There was no time-limit for the response. Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion.

@@

The individual trajectories reveal some points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By first averaging the x and y coordinates (in _cm_) of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. Both the horizontal and vertical distance of the reset point from the origin of the trajectory are converted to centimeters for purposes of model fitting (to keep them in the same unit). The vertical component can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.

Apart from reset points, we need a measure of illusion strength. We take a trajectory sample halfway between the starting point and the detected reset point. We then use the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength. This is comparable to $K \cdot V_i/V_e$ in Eq. 1, even when expressed in degrees angle.

@@

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **Left:** One participant's trajectories with very clear 'return' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear. **Middle:** Average reset point locations for all 9 participants in all 6 conditions (54 points) with standard deviations. (Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 2 (bottom right); the arctangent of the slope (internal over external velocity). Predictions from Eq. 1: Gray dashed line; $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.49$."}
#plotExampleData()
plotData()
```

@@

We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019]. A line through the origin with slope 0.49 ($K$ from Eq. 1) predicts illusion strength fairly well in this data, which is weaker than earlier ($K$=0.81) or with online tracking ($K$=0.72). We can also see that illusion strength measured with re-tracing is more variable. While the illusion is less strong and more variable, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

We test this with three simple models, based on straighforward trigonometry. First, we calculate the slope of a line ($s$) through the origin and a sample halfway to the reset point, indicating the strength of the illusion. If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict spatial offset from the physical path at the time of the reset, $x$:

@@

Where $L_y$ and $L_x$ are a temporal and spatial limit that are free parameters, and $s$ corresponds to the calculated slope. If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters for the single limit models indicate a temporal limit of ~1.4 _s_ (AIC=-26.2) and a spatial limit of ~1.8 _cm_ (AIC=-35.1), as illustrated in Figure 6 (left and middle). This is close to the mean values of the respective coordinates of reset points (Fig. 5, middle).  The model with two limits uses a temporal limit of 0.33 _s_ and a spatial limit of 1.47 _cm_ (AIC=-34.5). The model with a single spatial limit fits the data better than the model with a temporal limit ( _p_=.01, relative log-likelihood based on AICs) but is not better than the model with combined limits ( _p_=.73). This indicates that a spatial limit explains most of the data.

@@

Recently, Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion. In their model, when the position signals are weak, the motion signals bias perceived position away from the true location. In their case, this deviation saturated after about 200 _ms_ and the perceived path then continued parallel to the physical path, as in the “hit the wall” resets we described above [@Kwon2015, Fig. 3]. In contrast to their model, our participants in Experiment 2 showed quite variable timing for the reset point where the accumulation of offset ended. The mean time of these reset points was not 200 _ms_ but 1.5 _s_ and, more importantly, ranged from 0.5 _s_ to almost 4 _s_ (see Fig. 4). Moreover, 4/9 of our participants regularly reported “jump resets” that the Kwon et al. [-@Kwon2015] model cannot predict. The jump resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location. At that point, the build-up of the illusion may start again.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in this experiment also retraced the perceived path and reported "jump" resets. Nakayama and Holcombe also suggest that planned eye movements can trigger a “jump” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the jump resets that participants reported. However, "jump" resets occured at a fairly consistent location, which suggests that they are triggered by the stimulus and not by attentional distraction or eye movements.

>>>>

If $K$ is 0, then there is no illusion and the perceived direction matches the external motion direction. However, if $K$ is 1, then the internal motion has equal contribution with the external motion in determining the perceived direction. For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in both of our two experiments.

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **Left:** Individual trajectories. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity. **Top right:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading. **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path. In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$. Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$. Purple: best fit to our data, $K=0.72$."}
plotBoundedTracking()
```

@@

The illusion cannot drift continuously away from the physical path forever. It can be disrupted by a temporal break [@Lisi2015] or by distracting attention [@Nakayama2020]. These disruptions may cause the perceived path to stay at a fixed offset, travelling parallel to the physical path, or it may return toward the veridical position before resuming the illusory direction (Fig. 3). These resets may also occur spontaneously once the accumulation has gone on too long or too far. The purpose of this experiment is to capture this spontaneous resetting and determine its source.

@@

```{r fig3, echo=F, fig.height=4.55, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="\\label{fig:fig3}**Figure 3:** Double-Drift Resets. **Left:** At some point, the perceived position may stop moving further away from its actual path. For some people this takes the form of a ‘return’ reset, back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two. **Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinatee, the spatial offset from the vertical path. **Bottom right:** if they occur instead after some period of time, resets should have a constant Y coordinate, the time since the gabor’s motion began."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

@@

In both kinds of trials, participants first had to move the stylus to the start position of the gabor, and then fixate a point to the left or right of the centre of the workspace. The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass. After the gabor had disappeared, participants could respond in one of two ways. When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line. Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE. There was no time-limit for the response. Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, if there are any.

@@

The individual trajectories reveal some points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By first averaging the x and y coordinates (in _cm_) of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. Both the horizontal and vertical (straight ahead) distance of the reset point from the origin of the trajectory are converted to centimeters, and trajectories are scaled in both dimensions so that vertical distance from the origin is 13.5 cm. The vertical component can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.

Apart from reset points, we need a measure of illusion strength. We take a trajectory sample halfway between the starting point and the detected reset point. We then use the angular difference between the gabor's real trajectory and a straight line drawn through that point and the start of the trajectory as a measure of illusion strength.

@@

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **Left:** One participant's trajectories with very clear 'return' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear. **Middle:** Average reset point locations for all 9 participants in all 6 conditions (54 points) with standard deviations. (Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities. Blue dots from 3 s passes, red from 4 s passes, open circles are group averages. Predictions from Eq. 1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$."}
#plotExampleData()
plotData()
```

@@

We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019]. A line through the origin with slope 0.51 ($K$ from Eq. 1) predicts illusion strength fairly well in these data, which is weaker than the $K$=0.81 found by Cavanagh & Tse [-@Cavanagh2019] and lower than the $K$=0.72 found in Experiment 1 with online tracking. We can also see that illusion strength measured with re-tracing is more variable. While the illusion is less strong and more variable, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

We test this with three simple models, based on straighforward trigonometry. First, we calculate the slope of a line ($s$, time/illusory offset) through the origin and a sample halfway to the reset point, indicating the strength of the illusion. If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict spatial offset from the physical path at the time of the reset, $x$:

@@

Where $L_y$ and $L_x$ are a temporal and spatial limit that are free parameters, and $s$ corresponds to the calculated slope from the start of the trajectory to the halfway point. If we fit these simple models by minimizing the mean square error (errors are the absolute distance in centimeters between predicted and actual reset points), the optimal parameters for the single limit models indicate a temporal limit of ~1.82 _s_ (Eq. 2; AIC=-20.3) and a spatial limit of ~2.46 _cm_ (Eq. 3; AIC=-28.4), as illustrated in Figure 6 (left and middle). This is close to, but not the same as, the mean values of the respective coordinates of reset points (Fig. 5, middle). The model with two limits uses a temporal limit of 0.55 _s_ and a spatial limit of 1.89 _cm_ (Eq. 4a,b; AIC=-28.8). The model with a single spatial limit fits the data better than the model with a temporal limit ( _p_=.016, relative log-likelihood based on AICs) but is not better than the model with combined limits ( _p_=.87). This indicates that a spatial limit explains most of the data.

@@

Recently, Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion. In their model, when the position signals are weak, the motion signals bias perceived position away from the true location. In their case, this deviation saturated after about 200 _ms_ and the perceived path then continued parallel to the physical path, as in the “hit the wall” resets we described above [@Kwon2015, Fig. 3]. In contrast to their model, our participants in Experiment 2 showed quite variable timing for the reset point where the accumulation of offset ended. The resets were more stable in terms of spatial offset (Fig. 4, middle) but much more variable in terms of time from the start of the gabor's motion. In any case, the temporal mean of these reset points was not 200 _ms_ but 1.5 _s_ and ranged from 0.5 _s_ to almost 4 _s_ (see Fig. 4). Moreover, 4/9 of our participants regularly reported “return” resets that the Kwon et al. [-@Kwon2015] model does not predict. The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location. At that point, the build-up of the illusion may start again.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in this experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe also suggest that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.
